Developing a Visual Studio Code Language Support Extension for the Snail Programming Language

### 2023-06-15 paper.tex
feat: polish section 1-4

This paper will outline the process of building the Snail Language Support VS Code extension. Specificaly, we will introduce the structure of a VS Code extension that is meant to add support for new programming langauges. We will also discuss how this extension uses VS Code's language server protocol (LSP) to provide realtime error diagnostics \cite{Microsoft_2022a}. We will address how the Snail Language Support extension may be further developed to include debugging support with breakpoints, step in/out behavior, and variable inspection, particularly highlighting the rolw of VS Code's debug adapter protocol DAP \cite{Microsoft_2021}. Finally, we will review good software development practices such as version control and documentation.

@@

\begin{wrapfigure}{c}{0.3\textwidth}
    \begin{center}
        \includegraphics[height=0.3\textheight]{directory-structure.png}
        \caption{The typical directory structure of a VSCode language support extension}
        \label{fig:directory-structure}
    \end{center}
\end{wrapfigure}

@@

To inspect Snail Language Support's \lstinline{extension.ts} file more closely, see Appendix \ref{app:extension-ts}

@@

Next, and very importantly, we define where the source code is located for our extension. Through the \lstinline{main} attribute, we define the entry point for our extension. This is the file that contains the \lstinline{activate} function. For Snail Language Support, this is the \lstinline{/out/extension.js} file. Notice that this is the transpiled version of our \lstinline{src/extension.ts} file. For a more detailed look at Snail Language Support's \lstinline{package.json} file, see Appendix \ref{app:package-json}.

@@

Snail Language Support also provides language configuration settings supported by a VS Code extension. These configurations are found in \lstinline{language-configuration.json} file. We can define what is considered a commend through the \lstinline{comments.lineComment} or \lstinline{comments.blockComment} attributes, which allows a user to comment lines using VS Code's comment shortcut. Similarly, we can also define auto-closing structures such as parentheses or brackets through the \lstinline{autoClosingPairs} attribute, which allows VS Code to automatically place closing parentheses or brackets. This is particularly useful for snail, as more opening parentheses or brackets require a closing partner. By defining \lstinline{autoClosingPairs}, a developer can expend less mental energy ensuring that each parenthesis and bracket has a closing pair, and more mental energy on developing code. The \lstinline{language-configuration.json} file for Snail Language Support can be found in Appendix \ref{app:language-configuration}. For more attributes and configuration options, see the VS Code documentation on language configurations \cite{Microsoft_2023f}.

@@

Snail Language Support provides autocompletion suppport for common structures such as if-else clauses, while loops, and class definitions. This autocompletion is implemented through VS Code snippets. A snippet is a commonly used code structure that is defined by a \lstinline{prefix} (or a few options for a \lstinline{prefix}) and a \lstinline{body}. When a user types a snippet \lstinline{prefix}, they are given the option to replace the keyword with the \lstinline{body} of the snippet to complete the structure. Snippet bodies can also contain placeholders, which a user can visit in sequence while pressing their \lstinline{tab} button. For a full listing of the snippets that Snail Language Support provides, see Appendix \ref{app:snippets-json}. Similarly, for more information on snippets in VS Code, see the documentation \cite{Microsoft_2023g}.

@@

Snail Language Support also supports syntax highlighting for the snail language. VS Code's syntax highlighting makes heavy use of TextMate language grammars. A generic language grammar is a set of rules that govern what constitutes a valid statement for a given programming language \cite{pgrandinetti_2019}. A TextMate grammar is a specific format for defining a language grammar \cite{MacroMates_2021a}. TextMate grammars define patterns that match special elements of a text document and assign scope names to these elements. For example, keywords of a particular language might be assigned the \lstinline{keyword.control} scope name. These patterns are defined with regular expressions.

@@

Snail Language Support defines a TextMate grammar in the \lstinline{snail.tmLanguage.json} file, shown in Appendix \ref{app:snail-tmLanguage}. Taking a closer look at the structure of this document, we see three main relevant attributes: \lstinline{$schema}, \lstinline{patterns}, and \lstinline{repository}.

@@

Snail Language Support also features automatic lexing and parsing error diagnostics. In order to support this feature, Snail Language Support uses a language server that communicates via VS Code's langauge server protocol (LSP). We now discuss what a language server is, how it relates to the LSP, and how Snail Language Support uses langauge servers.

\subsection{What is a language server?}

\begin{wrapfigure}{r}{0.3\textwidth}
    \begin{center}
        \includegraphics[width=0.3\textwidth]{language-server.png}
        \caption{A development tool will communicate with a language server to display advanced diagnostics. The language server protocol standardizes this communication.}
        \label{fig:language-server}
    \end{center}
\end{wrapfigure}

A language server is a tool used to provide language-specific editing features such as advanced autocomplete, go-to definition, or automatic error checking \cite{Microsoft_2022a}. Many IDEs today utilize language servers. A development tool will launch a separate language server process to run in the background while a developer writes code. The development tool will then send a \emph{request} to the language server for advanced diagnostics, and the language server will \emph{respond} with the information requested by the development tool. The development tool will then display the diagnostics to the developer. This process is displayed in figure \ref{fig:language-server}.

Many language servers are implemented in the language they assist with. For example, a typical Python language server would be implemented in Python. Further, a single language server might have to interact with a number of different development tools. For example, a Python language server might have to interact with VS Code, PyCharm, and Spyder. Without a standard framework for communication between development tools and language servers, development tools and language servers must implement specific functionality for each pairing. This problem is displayed in Figure \ref{fig:no-lsp-theory}.

Each development tool has different standards and APIs for how they interact with language servers. To address this, a development tool may need to adjust how they request and recieve diagnostics from a language server. A multi-language tool might need to make these adjustments for every language server it interacts with. Alternatively, a language server might adjust how it recieves and responds to diagnostic requests for each development tool it interacts with. Neither of these options are ideal, and both options result in unnecessary duplicated logic.

@@

The Langauge Server Protocol (LSP) defines a standard framework for development tool and language server communication \cite{Microsoft_2022a}. This way, a development tool simply sends and recieves information according to the LSP standard, and is able to communicate with any language-specific language server this way. Similarly, a langauge server will be able to communicate with any development tool by communicating via the LSP standard.

@@

VS Code's LSP relies on JSON-RPC requests and responses \cite{Microsoft_2023i}. A request is a message that a development tool sends to a language server. This message contains an identifier, method to be executed, and paramaters to be provided to that method. In order to follow the LSP, a language server \textbf{must} send the corresponding response message upon recieving any requests. This response message contains the request identifier, the result of the executed method, and an error object if the request fails. For example, a development tool might send the \lstinline{Initialize} request to a language server, and a language server must respond with the \lstinline{InitializeResponse} before the development tool is allowed to send any further requests. Language servers are also allowed to send requests to development tools, and development tools are still required to respond with the corresponding response.

VS Code's LSP also contains notifications. A notification is a message that does not require a response. A development tool can send a notification to a language server and the language server can process the notification as it sees fit. Often, the language server will send information back to the development tool (in the form of another notification), but it is important to note that the language server is not required to respond.

With an established standard for communication, both development tools and language servers can streamline their communications systems. Additionally, a developer producing one of these tools or features can rely on a consistent standard to make the development process easier. This benefit is also illustrated in Figure \ref{fig:lsp-theory}.

@@

In the Snail Language Support extension, we use a language server to validate snail files before runtime, checking for lexer and parser errors. With the extension enabled, VS Code is able to display lexer and parser errors, and highlight the relevant location of the error, before a developer runs the program themselves. This behavior is illustrated in Figure \ref{fig:dev-diagnostics}.

\begin{figure}
    \begin{subfigure}{0.45\textwidth}
        \begin{center}
            \includegraphics[width=\textwidth]{dev-no-diagnostics.png}
            \subcaption{Developing snail code without realtime error diagnostics can make it difficult to spot syntax errors.}
            \label{fig:dev-no-diagnostics}
        \end{center}
    \end{subfigure}
    \hspace{0.05\textwidth}
    \begin{subfigure}{0.45\textwidth}
        \begin{center}
            \includegraphics[width=\textwidth]{dev-with-diagnostics.png}
            \subcaption{With Snail Language Support, a developer can spot easily spot syntax errors and fix them before runtime.}
            \label{fig:dev-with-diagnostics}
        \end{center}
    \end{subfigure}
    \caption{Demonstrating the benefit of realtime error diagnostics provided in Snail Language Support.}
    \label{fig:dev-diagnostics}
\end{figure}

@@

@@

\begin{wrapfigure}{C}{0.3\textwidth}
    \includegraphics[width=0.3\textwidth]{lsp-sequence.png}
    \caption{Modelling the interactions between a developer, a VS Code client, and snail language server while a developer opens and modifies a snail file.}
    \label{fig:lsp-sequence}
\end{wrapfigure}

A high-level sequence of interactions between a developer, VS Code, and the snail language server is documented in Figure \ref{fig:lsp-sequence}. First, a user opens a program written in snail (1). When VS Code detects this activity, it launches the snail language server (2). Next, VS Code provides the snail program for the language server to verify (3) and the language server executes the snail program, using the aforementioned \lstinline{snail -s} option (4). The snail language server returns the error diagnostics to VS Code (5), and VS code displays those diagnostics to the developer (6). The developer further modifies the snail file, and VS Code and the snail language server continue to communicate via LSP (7-10). When a developer is finished, they can close the snail file (11), and the VS Code client shuts down the snail language server (12).

>>>>

This paper will outline the process of building the Snail Language Support VS Code extension.\footnote{This paper is formatted using \LaTeX. Recognizing that using \LaTeX \hspace{1pt} is really hard, the \LaTeX \hspace{1pt} source code for this paper can be found hosted on GitHub: \url{https://github.com/CharlesReinhardt/snail-language-support-paper/}} Specificaly, we will introduce the structure of a VS Code extension that is meant to add support for new programming langauges. We will also discuss how this extension uses VS Code's language server protocol (LSP) to provide realtime error diagnostics \cite{Microsoft_2022a}. We will address how the Snail Language Support extension may be further developed to include debugging support with breakpoints, step in/out behavior, and variable inspection, particularly highlighting the role of VS Code's debug adapter protocol (DAP) \cite{Microsoft_2021}. Finally, we will review good software development practices such as version control and documentation.

@@

\begin{wrapfigure}{R}{0.3\textwidth}
    \begin{center}
        \includegraphics[height=0.3\textheight]{directory-structure.png}
        \caption{
            \centering 
                The typical directory structure of a VSCode language support extension.
        }
        \label{fig:directory-structure}
    \end{center}
\end{wrapfigure}

@@

To inspect Snail Language Support's \lstinline{extension.ts} file more closely, see appendix \ref{app:extension-ts}

@@

Next, and very importantly, we define where the source code is located for our extension. Through the \lstinline{main} attribute, we define the entry point for our extension. This is the file that contains the \lstinline{activate} function. For Snail Language Support, this is the \lstinline{/out/extension.js} file. Notice that this is the transpiled version of our \lstinline{src/extension.ts} file. For a more detailed look at Snail Language Support's \lstinline{package.json} file, see appendix \ref{app:package-json}.

@@

Snail Language Support also provides language configuration settings supported by a VS Code extension. These configurations are found in \lstinline{language-configuration.json} file. We can define what is considered a commend through the \lstinline{comments.lineComment} or \lstinline{comments.blockComment} attributes, which allows a user to comment lines using VS Code's comment shortcut. Similarly, we can also define auto-closing structures such as parentheses or brackets through the \lstinline{autoClosingPairs} attribute, which allows VS Code to automatically place closing parentheses or brackets. This is particularly useful for snail, as more opening parentheses or brackets require a closing partner. By defining \lstinline{autoClosingPairs}, a developer can expend less mental energy ensuring that each parenthesis and bracket has a closing pair, and more mental energy on developing code. The \lstinline{language-configuration.json} file for Snail Language Support can be found in appendix \ref{app:language-configuration}. For more attributes and configuration options, see the VS Code documentation on language configurations \cite{Microsoft_2023f}.

@@

Snail Language Support provides autocompletion suppport for common structures such as if-else clauses, while loops, and class definitions. This autocompletion is implemented through VS Code snippets. A snippet is a commonly used code structure that is defined by a \lstinline{prefix} (or a few options for a \lstinline{prefix}) and a \lstinline{body}. When a user types a snippet \lstinline{prefix}, they are given the option to replace the keyword with the \lstinline{body} of the snippet to complete the structure. Snippet bodies can also contain placeholders, which a user can visit in sequence while pressing their \lstinline{tab} button. For a full listing of the snippets that Snail Language Support provides, see appendix \ref{app:snippets-json}. Similarly, for more information on snippets in VS Code, see the documentation \cite{Microsoft_2023g}.

@@

Snail Language Support also supports syntax highlighting for the snail language as shown in figure \ref{syntax-highlight}. VS Code's syntax highlighting makes heavy use of TextMate language grammars. A generic language grammar is a set of rules that govern what constitutes a valid statement for a given programming language \cite{pgrandinetti_2019}. A TextMate grammar is a specific format for defining a language grammar \cite{MacroMates_2021a}. TextMate grammars define patterns that match special elements of a text document and assign scope names to these elements. For example, keywords of a particular language might be assigned the \lstinline{keyword.control} scope name. These patterns are defined with regular expressions.

\begin{figure}[b]
    \begin{subfigure}[c][][c]{0.45\textwidth}
        \begin{center}
            \includegraphics[width=\textwidth]{dev-no-highlight-light.png}
            \caption{Developing snail code without syntax highlighting can be a mentally taxing experience.}
            \label{fig:dev-no-highlight}
        \end{center}
    \end{subfigure}
    \hspace{0.05\textwidth}
    \begin{subfigure}[c][][c]{0.45\textwidth}
        \begin{center}
            \includegraphics[width=\textwidth]{dev-with-highlight-light.png}
            \caption{Syntax highlighting makes developing snail code much easier.}
            \label{fig:dev-with-highlight}
        \end{center}
    \end{subfigure}
    \caption{Demonatrating the benefit of Snail Language Support's syntax highlighting feature.}
    \label{fig:syntax-highlight}
\end{figure}

@@

Snail Language Support defines a TextMate grammar in the \lstinline{snail.tmLanguage.json} file, shown in appendix \ref{app:snail-tmLanguage}. Taking a closer look at the structure of this document, we see three main relevant attributes: \lstinline{$schema}, \lstinline{patterns}, and \lstinline{repository}.

@@

Snail Language Support also provides realtime lexing and parsing error diagnostics when developing code in snail. To do this, Snail Language Support uses a language server that communicates via VS Code's langauge server protocol (LSP). We now discuss what a language server is, how it relates to the LSP, and how Snail Language Support uses langauge servers.

\subsection{What is a Language Server?}

\begin{wrapfigure}{r}{0.5\textwidth}
    \begin{center}
        \includegraphics[width=0.5\textwidth]{language-server.png}
        \caption{
            \centering 
            A development tool will communicate with a language server to display advanced diagnostics.
        }
        \label{fig:language-server}
    \end{center}
\end{wrapfigure}

A language server is a tool used to provide language-specific editing features such as advanced autocomplete, go-to definition, or automatic error checking \cite{Microsoft_2022a}. Many IDEs today utilize language servers by launching a language server process that runs in the background while a developer writes code. The development tool then sends a \emph{requests} to the language server for, and the language server \emph{responds} with the information requested by the development tool. The development tool then displays the information to the developer. This process is displayed in figure \ref{fig:language-server}.

Many language servers are implemented in the language they assist with \cite{Microsoft_2023j}. For example, a typical Python language server would be implemented in Python. This makes integrating language servers into development tools difficult, as many development tools are written in languages other than the language they are meant to develop. Further, a single language server might have to interact with a number of different development tools. For example, a Python language server might have to interact with VS Code, PyCharm, and Spyder. Without a standard framework for communication between development tools and language servers, development tools and language servers must implement specific functionality for each pairing. This problem is displayed in figure \ref{fig:no-lsp-theory}.

Each development tool has different standards for how they interact with language servers. To address this, a development tool may need to adjust how they request and recieve information from each language server it interacts with. Alternatively, a language server might adjust how it recieves and responds to requests for each development tool it interacts with. Both options are not ideal.

@@

The Langauge Server Protocol (LSP) defines a standard framework for development tool and language server communication \cite{Microsoft_2022a}. A development tool is able to communicate with any language-specific language server by sending requests and recieving responses according to the LSP standard.

@@

VS Code's LSP relies on requests and responses \cite{Microsoft_2023i}. A request is a message that a development tool sends to a language server, asking the language server to perform some operation. A language server responds to the request by sending a corresponding response message with information about the executed operation. In order to follow the LSP, a language server \textbf{must} send a corresponding response upon recieving any requests. For example, a development tool might send the \lstinline{Initialize} request to a language server, and a language server must respond with the \lstinline{InitializeResponse} before the development tool is allowed to send any further requests. This contract goes both ways: when a language server sends a request to a developement tool, the development tool must respond.

VS Code's LSP also uses notifications. A notification is a message that does not require a response. A development tool can send a notification to a language server and the language server can process the notification as it sees fit. Often, the language server will send information back to the development tool (in the form of another notification), but it is important to note that the language server is not required to respond.

With an established standard for communication as shown in \ref{fig:lsp-theory}, both development tools and language servers can streamline their communications systems. Additionally, a developer producing one of these tools or features can rely on a consistent standard to make the development process easier.

@@

In the Snail Language Support extension, we use a language server to validate snail files before runtime, checking for lexer and parser errors. With the extension enabled, VS Code is able to display lexer and parser errors, and highlight the relevant location of the error, before a developer runs the program themselves. This behavior is illustrated in figure \ref{fig:dev-diagnostics}.

\begin{figure}
    \begin{subfigure}{0.45\textwidth}
        \begin{center}
            \includegraphics[width=\textwidth]{dev-no-diagnostics-light.png}
            \subcaption{
                \centering
                    Developing snail code without realtime error diagnostics can make it difficult to spot syntax errors.
                }
            \label{fig:dev-no-diagnostics}
        \end{center}
    \end{subfigure}
    \hspace{0.05\textwidth}
    \begin{subfigure}{0.45\textwidth}
        \begin{center}
            \includegraphics[width=\textwidth]{dev-with-diagnostics-light.png}
            \subcaption{
                \centering
                    With Snail Language Support, a developer can spot easily spot syntax errors and fix them before runtime.
                }
            \label{fig:dev-with-diagnostics}
        \end{center}
    \end{subfigure}
    \caption{
        \centering
            Demonstrating the benefit of realtime error diagnostics provided in Snail Language Support.
        }
    \label{fig:dev-diagnostics}
\end{figure}

@@

\begin{wrapfigure}{R}{0.5\textwidth}
    \includegraphics[width=0.5\textwidth]{lsp-sequence.png}
    \caption{
        \centering
            Modelling the interactions between a developer, a VS Code client, and snail language server while a developer opens and modifies a snail file.
    }
    \label{fig:lsp-sequence}
\end{wrapfigure}

@@

A high-level sequence of interactions between a developer, VS Code client, and snail language server is documented in figure \ref{fig:lsp-sequence}. First, a user opens a program written in snail (1). When the VS Code client detects this activity, it launches a snail language server (2). Next, the VS Code client provides the snail program to verify (3) and the language server executes the snail program, using the aforementioned \lstinline{snail -s} option (4). The snail language server returns the error diagnostics to the VS Code client (5) and the client displays these diagnostics to the developer (6). While the developer further modifies the snail file, the VS Code client and snail language server continue to communicate via LSP (7-10). When a developer is finished, they can close the snail file (11) and the VS Code client shuts down the snail language server (12).
